XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-27 talk/icooolps2011/paper.tex
some changes, some comments

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand{\commentout}[1]{}

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, implementing just an interpreter is already an arduous
task. Implementing them efficiently with a just-in-time compiler is
nigh-impossible, because or their many corner-cases.

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea using the different technique of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they implement a dynamic language in some
implementation language. In addition they build a tracing JIT for that implementation
language. The tracing JIT then traces through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project which used the for dynamic
assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

Because the traces always correspond to a concrete execution they cannot
contain any control flow splits. Therefore they encode the control flow
decisions needed to stay on the trace with the help of \emph{guards}. Those are
operations that check that the assumptions are still true when the trace is
later executed with different values.

@@

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Therefore, we give every class a version number, which is changed every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

>>>>

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\pedronis[1]{\nb{PEDRONIS}{#1}}
\newcommand{\commentout}[1]{}

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, implementing just an interpreter is already an arduous
\pedronis{XXX ardous seems a bit too strong} task. Implementing them efficiently with a just-in-time compiler (JIT) is
extremely challenging, because or their many corner-cases.

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the implementation
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT then will trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project which used them for dynamic
assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

Because the traces always correspond to a concrete execution they cannot
contain any control flow splits. Therefore they encode the control flow
decisions needed to stay on the trace with the help of \emph{guards}. Those are
operations that check that the assumptions are still true when the compiled trace is later executed with different values.

@@

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
\pedronis{XXX style: do we want to use you, or we and one} see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}
